---
title: "Chapter 8"
subtitle: "Introduction to Multilevel Models"
output:
  pdf_document:
    number_sections: yes
  html_document: default
---

```{r,include=F}
if(knitr::is_html_output()){options(knitr.table.format = "html")} else {options(knitr.table.format = "latex")}
```

# Two Level Longitudinal Data {#ch-lon}

## Learning objectives

  After finishing this chapter, you should be able to:

- Recognize longitudinal data as a special case of multilevel data, with time at Level One.
- Consider patterns of missingness and implications of that missing data on multilevel analyses.
- Apply exploratory data analysis techniques specific to longitudinal data.
- Build and understand a taxonomy of models for longitudinal data.
- Interpret model parameters in multilevel models with time at Level One.
- Compare models, both nested and not, with appropriate statistical tests and summary statistics.
- Consider different ways of modeling the variance-covariance structure in longitudinal data.

```{r load_packages9, message = FALSE}
# Packages required for Chapter 9
library(GGally)
library(data.table)
library(Hmisc)
library(mice)
library(lattice)
library(nlme)
library(reshape2)
library(MASS)
library(mnormt)
library(lme4)
library(gridExtra) 
library(knitr)
library(kableExtra)
library(tidyverse)
```

Here's a reference @BBCNEWS1995

and another with brackets [@Bayer2011]

and another with brackets and minus sign [-@Allison2005]

does double brackets help with caps: @idre2018

Also @Bayer2011, @Awad2017, @Dahlquist2011, @Strutz2013, @Fast2011, @Trinh2018, @Renier2018, @Angell2010, @Pray2009